A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-23 tex/codatadp.tex
more

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is $\thD$-unsatisfiable.
\end{theorem}
\begin{proof}
The proof is by structural induction on the derivation tree with root node $\Ec$.
If the tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, the construction of $\Val$ indicates that the class of values that term~$t$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees
whose root nodes are the result of applying the rule \rn{Split} on $t\typ{\delta}$,
by the induction hypothesis $\Ec \mathrel\cup t \teq \smash{\const C_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^{\,\smash{n_{\negvthinspace j}}}_{\negvthinspace j}( t ) \bigr)}$ is unsatisfiable
for each $\const C_{\negvthinspace j} \in \smash{\Ctr^\delta}$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq \const C_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^{\,\smash{n_{\negvthinspace j}}}_{\negvthinspace j}( t ) \bigr)$,
this means $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree
whose root node is $\Ec \mathrel\cup t \teq u$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Unique}, or \rn{Single}.
In all cases, we have that $\Ec \models_{\thD} t \teq u$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Unique}, our construction of $\Val$ indicates that the value of $t$ and $u$ are equivalent in all models of $\Ec$.
For \rn{Single}, $t$~and~$u$ must have the same value since the cardinality of their type is $1$.
By the induction hypothesis, $\Ec \mathrel\cup t \teq u$ is $\thD$-unsatisfiable
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a strict subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its strict subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
In particular, note that their arguments have the same expansion with respect to $t$,
that is,
$\interp{t}{\MU y.\; \const{C}( y )} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\smash{\{y\}}} ) =$
$\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\smash{\{y\}}} ) =$
$\MU y.\; \const{C}( y )$.
The term $t$ of the form $\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
%
\begin{align*}
\interp{t}{\MU y.\; \const{C}( x )}
& = \MU y.\; \const{C}( \interp{t}{x}^{\smash{\{y\}}} ) \\
& = \MU y.\; \const{C}( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\smash{\{y\}}} ) \\
& = \MU y.\; \const{C}( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\smash{\{x,y\}}} ) ) \\
& = \MU y.\allowbreak\; \const{C}( \MU x.\allowbreak\; \const{C}( y ) )
\end{align*}
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct its \emph{normal form} $\nf{t}$
by replacing all of the self-similar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

\begin{report}
\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim \interp{t}{\vec w}$.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

\begin{lemma}
\label{lem:mu-norm-interp}
If $t$ is normal and has subterm $u$, then $\interp{t}{u}$ is normal.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

We now define the class of interpretations for $\thD$.
$\M( \tau )$ denotes the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t$, $\M( t )$ denotes the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
\emph{All} types are interpreted as a set of $\mu$-terms.
This is done to simplify the presentation.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For example, the $\mu$-term $\MU x.\; \const{0}$ representing the integer constant zero
is simply $\const{0}$ after dropping the $\mu$-binding of $x$.

>>>>

\begin{theorem}[Refutation Soundness]%
\label{thm:rs}%
\afterDot
If there exists a closed derivation tree with root node\/ $\Ec$, then\/ $\Ec$ is $\thD$-unsatisfiable.
\end{theorem}
\begin{proof}
The proof is by structural induction on the derivation tree with root node $\Ec$.
If the tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, the construction of $\Val$ indicates that the class of values that term~$t$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the child nodes of $\Ec$ are closed derivation trees
whose root nodes are the result of applying the rule \rn{Split} on $t\typ{\delta}$,
by the induction hypothesis $\Ec \mathrel\cup t \teq \smash{\const C_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^{\,\smash{n_{\negvthinspace j}}}_{\negvthinspace j}( t ) \bigr)}$ is unsatisfiable
for each $\const C_{\negvthinspace j} \in \smash{\Ctr^\delta}$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq \const C_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^{\,\smash{n_{\negvthinspace j}}}_{\negvthinspace j}( t ) \bigr)$,
this means $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child node of $\Ec$ is a closed derivation tree
whose root node is $\Ec \mathrel\cup t \teq u$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Unique}, or \rn{Single}.
In all cases, we have that $\Ec \models_{\thD} t \teq u$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Unique}, our construction of $\Val$ indicates that the value of $t$ and $u$ are equivalent in all models of $\Ec$.
For \rn{Single}, $t$~and~$u$ must have the same value since the cardinality of their type is $1$.
By the induction hypothesis, $\Ec \mathrel\cup t \teq u$ is $\thD$-unsatisfiable
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

The $\mu$-term $u$
is a \emph{self-similar subterm} of~$t$ if
$u$ is a strict subterm of $t$,
$t$ is of the form $\MU x.\; \const{C}( t_1, \ldots, t_n )$,
$u$ it is of the form $\MU y.\; \const{C}( u_1, \ldots, u_n )$,
and $\interp{t}{ t_k } \aequiv \interp{t}{ u_k }$ for all $k \in [n]$.
The term $t$ is \emph{normal} if it does not contain a self-similar subterm
and all of its strict subterms are also normal.
For example, $t = \MU x.\; \const{C}( \MU y.\; \const{C}( y ) )$ is abnormal
because $\MU y.\; \const{C}( y )$ is a self-similar subterm of $t$.
In particular, note that their arguments have the same expansion with respect to $t$,
that is,
$\interp{t}{\MU y.\; \const{C}( y )} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\smash{\{y\}}} ) =$
$\MU y.\; \const{C}( y )$
is $\alpha$-equivalent to
$\interp{t}{y} =$
$\MU y.\; \const{C}( \interp{t}{y}^{\smash{\{y\}}} ) =$
$\MU y.\; \const{C}( y )$.
The term $t = \MU x.\; \const{C}( \MU y.\; \const{C}( x ) )$ is also abnormal,
since $\MU y.\; \const{C}( x )$ is a self-similar subterm of $t$,
noting that
%
\begin{align*}
\interp{t}{\MU y.\; \const{C}( x )}
& = \MU y.\; \const{C}( \interp{t}{x}^{\smash{\{y\}}} ) \\
& = \MU y.\; \const{C}( \interp{t}{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ) }^{\smash{\{y\}}} ) \\
& = \MU y.\; \const{C}( \MU x.\; \const{C}( \interp{t}{\MU y.\; \const{C}( x )}^{\smash{\{x,y\}}} ) ) \\
& = \MU y.\allowbreak\; \const{C}( \MU x.\allowbreak\; \const{C}( y ) )
\end{align*}
is $\alpha$-equivalent to $\interp{t}{x} = t$.
% for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct its \emph{normal form} $\nf{t}$
by replacing all of the self-similar subterms of $t$ with $x$
%and by replacing all of the abnormal subterms of $t$ with their corresponding normal forms.
and by normalizing the subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

\begin{report}
\begin{lemma}
\label{lem:mu-norm-arg}
If the $\mu$-terms $\vec u$ are in normal form and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim \interp{t}{\vec w}$\vthinspace.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

\begin{lemma}
\label{lem:mu-norm-interp}
If the $\mu$-term $t$ is normal and $u$ is a subterm of $t$, then $\interp{t}{u}$ is normal.
\end{lemma}
\begin{proof}
\rem{TODO}
\end{proof}

@@

We now define the class of interpretations for $\thD$.
$\M( \tau )$ denotes the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
For a term $t\typ{\tau}$, $\M( t ) \in \M( \tau )$ denotes the interpretation of $t$ in $\M$.
To simplify the presentation, all types are interpreted as sets of $\mu$-terms.
Since all values of types in $\Data$ and $\Nondata$ are acyclic,
a typical representation for these values can be obtained by
dropping each of its $\mu$-bindings. 
For example, the $\mu$-term $\MU x.\; \const{Z}$ representing the integer constant zero
is simply $\const{Z}$ after dropping the $\mu$-binding of $x$.
